---
title: "Recommender System"
author: Adeesh Naik
date: December 11th, 2021
output: html_document
---

```{r setup, include=FALSE}
  library(recommenderlab)
  library(ggplot2)                       
  library(data.table)
  library(reshape2)
  library(data.table)
  

```

## Business Problem

The objective is to develop a recommendation engine the recommends movie to users. It provides suggestions to the users through a filtering process that is based on user preferences and browsing history. 
The information about the user is taken as an input that is in the form of browsing data. This information reflects the prior viewing of the movies as well as the assigned ratings.

Recommendations is provided in respect with the other users who might have a similar viewing history or preferences. There are two types of recommendation systems – Content-Based Recommendation System and Collaborative Filtering Recommendation and we will be using collaborative filtering recommendation system and more specifically, ITEM based collaborative recommendation system (IBCF).

## Data Source

There are 2 sources of data in the form of csv :
1. movie_data.csv  - A list of movies with Unique ID and their genres (can be multiple genres for same  movies)
2. rating_data.csv - A database of users and their rating for multiple movies in the above data set.

```{r 1, echo=TRUE}
  setwd("C:/Users/Naika/Desktop/Personal Space/Data Analytics/ANA 515 - Data Storage/Week 8/IMDB-Dataset")  
  movie_data <- read.csv("movies.csv",stringsAsFactors=FALSE)
  rating_data <- read.csv("ratings.csv")
  
  str(movie_data)
  str(rating_data)
  
```

## Data Exploration & Cleanup

From the above table, we observe that the userId column, as well as the movieId column, consist of integers. 
There are `r nrow(movie_data)` number of movies in the movie data set. The rating data contains `r nrow(rating_data)`

```{r 2, echo=FALSE}
  summary(movie_data)   
  head(movie_data)
  
  summary(rating_data)   
  head(rating_data)
```
  
 Since there can be multiple genres for each movie,  we need to convert them into a more usable format. In order to do so, we will first create a matrix that comprises of corresponding genres for each of the movies. 
 
```{r 3, echo=FALSE} 
  movie_genre <- as.data.frame(movie_data$genres, stringsAsFactors=FALSE)
  
  movie_genre2 <- as.data.frame(tstrsplit(movie_genre[,1], '[|]', 
                                          type.convert=TRUE), 
                                stringsAsFactors=FALSE) 
  colnames(movie_genre2) <- c(1:10)
  list_genre <- c("Action", "Adventure", "Animation", "Children", 
                  "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                  "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                  "Sci-Fi", "Thriller", "War", "Western")
  genre_mat1 <- matrix(0,10330,18)
  genre_mat1[1,] <- list_genre
  colnames(genre_mat1) <- list_genre
  for (index in 1:nrow(movie_genre2)) {
    for (col in 1:ncol(movie_genre2)) {
      gen_col = which(genre_mat1[1,] == movie_genre2[index,col]) 
      genre_mat1[index+1,gen_col] <- 1
    }
  }
  genre_mat2 <- as.data.frame(genre_mat1[-1,], stringsAsFactors=FALSE) #remove first row, which was the genre list
  for (col in 1:ncol(genre_mat2)) {
    genre_mat2[,col] <- as.integer(genre_mat2[,col]) #convert from characters to integers
  } 
  str(genre_mat2)
```  


For the movie recommendation system to make sense of our ratings through recommenderlabs, we have to convert our matrix into a sparse matrix one. 
This new matrix is of the class ‘realRatingMatrix’. This is performed as follows:

```{r 5, echo=FALSE} 

ratingMatrix <- dcast(rating_data, userId~movieId, value.var = "rating", na.rm=FALSE)
ratingMatrix <- as.matrix(ratingMatrix[,-1]) #remove userIds
#Convert rating matrix into a recommenderlab sparse matrix
ratingMatrix <- as(ratingMatrix, "realRatingMatrix")
ratingMatrix

```

Collaborative Filtering involves suggesting movies to the users that are based on collecting preferences from many other users. For example, if a user A likes to watch action films and so does user B, then the movies that the user B will watch in the future will be recommended to A and vice-versa. Therefore, recommending movies is dependent on creating a relationship of similarity between the two users. With the help of recommenderlab, we can compute similarities using various operators like cosine, pearson as well as jaccard.

```{r 6, echo=FALSE}  

recommendation_model <- recommenderRegistry$get_entries(dataType = "realRatingMatrix")
names(recommendation_model)

lapply(recommendation_model, "[[", "description")

recommendation_model$IBCF_realRatingMatrix$parameters


similarity_mat <- similarity(ratingMatrix[1:4, ],
                             method = "cosine",
                             which = "users")
as.matrix(similarity_mat)
image(as.matrix(similarity_mat), main = "User's Similarities")

```

In the above matrix, each row and column represents a user. We have taken four users and each cell in this matrix represents the similarity that is shared between the two users.
Next, delineate the similarity that is shared between the films –

```{r 7, echo=FALSE}  

movie_similarity <- similarity(ratingMatrix[, 1:4], method =
                                 "cosine", which = "items")
as.matrix(movie_similarity)
image(as.matrix(movie_similarity), main = "Movies similarity")

```

Next, extract the most unique ratings and will create a table of ratings that will display the most unique ratings.

```{r 8, echo=FALSE}  
rating_values <- as.vector(ratingMatrix@data)
unique(rating_values) # extracting unique ratings
Table_of_Ratings <- table(rating_values) # creating a count of movie ratings
Table_of_Ratings
```

Explore the most viewed movies in our dataset. We will first count the number of views in a film and then organize them in a table that would group them in descending order.
 
```{r 9, echo=FALSE}  

library(ggplot2)
movie_views <- colCounts(ratingMatrix) # count views for each movie
table_views <- data.frame(movie = names(movie_views),
                          views = movie_views) # create dataframe of views
table_views <- table_views[order(table_views$views,
                                 decreasing = TRUE), ] # sort by number of views
table_views$title <- NA
for (index in 1:10325){
  table_views[index,3] <- as.character(subset(movie_data,
                                              movie_data$movieId == table_views[index,1])$title)
}
table_views[1:6,]
```


A bar plot for the total number of views of the top films using ggplot2.

```{r 10, echo=FALSE}  
ggplot(table_views[1:6, ], aes(x = title, y = views)) +
  geom_bar(stat="identity", fill = 'steelblue') +
  geom_text(aes(label=views), vjust=-0.3, size=3.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Total Views of the Top Films")
```

A heatmap of the movie ratings, this heatmap will contain first 25 rows and 25 columns as follows –

```{r 11, echo=FALSE}  
image(ratingMatrix[1:20, 1:25], axes = FALSE, main = "Heatmap of the first 25 rows and 25 columns")
```

## Data Preparation


Data preparation is done in the following three steps –

 - Selecting useful data.
 - Normalizing data.
 - Binarizing the data.
 
For finding useful data in our dataset, set the threshold for the minimum number of users who have rated a film as 50. This is also same for minimum number of views that are per film. This way, we have filtered a list of watched films from least-watched ones.

```{r 12, echo=FALSE}
movie_ratings <- ratingMatrix[rowCounts(ratingMatrix) > 50,
                              colCounts(ratingMatrix) > 50]
movie_ratings
```

From the above output of ‘movie_ratings’, we observe that there are 420 users and 447 films as opposed to the previous 668 users and 10325 films. We can now delineate our matrix of relevant users as follows –

```{r 13, echo=FALSE}
minimum_movies<- quantile(rowCounts(movie_ratings), 0.98)
minimum_users <- quantile(colCounts(movie_ratings), 0.98)
image(movie_ratings[rowCounts(movie_ratings) > minimum_movies,
                    colCounts(movie_ratings) > minimum_users],
      main = "Heatmap of the top users and movies")
```

Visualizing the distribution of the average ratings per user.

```{r 14, echo=FALSE}
average_ratings <- rowMeans(movie_ratings)
qplot(average_ratings, fill=I("steelblue"), col=I("red")) +
  ggtitle("Distribution of the average rating per user")
```

In the case of some users, there can be high ratings or low ratings provided to all of the watched films. This will act as a bias while implementing our model. In order to remove this, normalize the data. 
Normalization is a data preparation procedure to standardize the numerical values in a column to a common scale value. This is done in such a way that there is no distortion in the range of values. Normalization transforms the average value of our ratings column to 0. We then plot a heatmap that delineates our normalized ratings.

```{r 15, echo=FALSE}
normalized_ratings <- normalize(movie_ratings)
sum(rowMeans(normalized_ratings) > 0.00001)
image(normalized_ratings[rowCounts(normalized_ratings) > minimum_movies,
                         colCounts(normalized_ratings) > minimum_users],
      main = "Normalized Ratings of the Top Users")
```

In the final step of data preparation, binarize the data. Binarizing the data means that we have two discrete values 1 and 0, which will allow our recommendation systems to work more efficiently. We will define a matrix that will consist of 1 if the rating is above 3 and otherwise it will be 0.

```{r 16, echo=FALSE}
binary_minimum_movies <- quantile(rowCounts(movie_ratings), 0.95)
binary_minimum_users <- quantile(colCounts(movie_ratings), 0.95)
#movies_watched <- binarize(movie_ratings, minRating = 1)
good_rated_films <- binarize(movie_ratings, minRating = 3)
image(good_rated_films[rowCounts(movie_ratings) > binary_minimum_movies,
                       colCounts(movie_ratings) > binary_minimum_users],
      main = "Heatmap of the top users and movies")
```


## Modelling

We will develop our very own Item Based Collaborative Filtering System. This type of collaborative filtering finds similarity in the items based on the people’s ratings of them. The algorithm first builds a similar-items table of the customers who have purchased them into a combination of similar items. This is then fed into the recommendation system.

The similarity between single products and related products can be determined with the following algorithm –

For each Item i1 present in the product catalog, purchased by customer C.
And, for each item i2 also purchased by the customer C.
Create record that the customer purchased items i1 and i2.
Calculate the similarity between i1 and i2.


We will build this filtering system by splitting the dataset into 80% training set and 20% test set.

```{r 17, echo=FALSE}
sampled_data<- sample(x = c(TRUE, FALSE),
                      size = nrow(movie_ratings),
                      replace = TRUE,
                      prob = c(0.8, 0.2))
training_data <- movie_ratings[sampled_data, ]
testing_data <- movie_ratings[!sampled_data, ]
```

We will now explore the various parameters of our Item Based Collaborative Filter. These parameters are default in nature. In the first step, k denotes the number of items for computing their similarities. 
Here, k is equal to 30. Therefore, the algorithm will now identify the k most similar items and store their number. We use the cosine method which is the default.

```{r 18, echo=FALSE}
recommendation_system <- recommenderRegistry$get_entries(dataType ="realRatingMatrix")
recommendation_system$IBCF_realRatingMatrix$parameters


recommen_model <- Recommender(data = training_data,
                              method = "IBCF",
                              parameter = list(k = 30))
recommen_model
class(recommen_model)
```

Using the getModel() function, we will retrieve the recommen_model. We will then find the class and dimensions of our similarity matrix that is contained within model_info. 
Finally, we will generate a heatmap, that will contain the top 20 items and visualize the similarity shared between them.

```{r 19, echo=FALSE}
model_info <- getModel(recommen_model)
class(model_info$sim)
dim(model_info$sim)
top_items <- 20
image(model_info$sim[1:top_items, 1:top_items],
      main = "Heatmap of the first rows and columns")
```

In the next step , we will carry out the sum of rows and columns with the similarity of the objects above 0. We will visualize the sum of columns through a distribution as follows –

```{r 20, echo=FALSE}
sum_rows <- rowSums(model_info$sim > 0)
table(sum_rows)

sum_cols <- colSums(model_info$sim > 0)
qplot(sum_cols, fill=I("steelblue"), col=I("red"))+ ggtitle("Distribution of the column count")
```

We will create a top_recommendations variable which will be initialized to 10, specifying the number of films to each user. We will then use the predict() function that will identify similar items and will rank them appropriately. Here, each rating is used as a weight. Each weight is multiplied with related similarities.

```{r 21, echo=FALSE}
top_recommendations <- 10 # the number of items to recommend to each user
predicted_recommendations <- predict(object = recommen_model,
                                     newdata = testing_data,
                                     n = top_recommendations)
predicted_recommendations
```

Here are the recommendation for the first user using the recommendation engine that we just built

```{r 22, echo=FALSE}
user1 <- predicted_recommendations@items[[1]] # recommendation for the first user
movies_user1 <- predicted_recommendations@itemLabels[user1]
movies_user2 <- movies_user1
for (index in 1:10){
  movies_user2[index] <- as.character(subset(movie_data,
                                             movie_data$movieId == movies_user1[index])$title)
}
movies_user2
```




